Equations & R Code

### 2016-10-31 manuscript/coexistence-stability_ms.Rmd
editing and adding in PBA suggestions

\begin{abstract}
Theory relating species richness to ecosystem stability typically ignores interactions between environmental variability and species coexistence, or fluctuation-dependent coexistence.
This may explain why we lack general explanations for observed deviations from the predicted positive diversity-stability relationship.
It also limits our ability to predict the consequences of increasing environmental variability.
We use a consumer-resource model to explore how fluctuation-dependent coexistence via the storage effect and relative nonlinearity affects ecosystem stability.
We show that a negative, rather than positive, diversity-stability relationship is possible when ecosystem function is sampled across a natural diversity gradient.
We also describe how fluctuation-dependent coexistence can buffer ecosystem functioning against increasing environmental variability by allowing more species to coexist and thus contribute to portfolio effects.
Our work provides a general explanation for non-positive diversity-stability relalationships and highlights the importance of conserving regional species pools so that species can be added to fluctuation-dependent communities as environmental variability increases.
\vspace{2em}
\end{abstract}

@@

INTRODUCTION
------------
@MacArthur1955, @Elton1958, and even Darwin [@Turnbull2013] recognized that species can compensate for each other and stabilize functioning in ecosystems subject to temporal variation in environmental conditions. 
This idea underlies the "insurance hypothesis" [@Yachi1999], which suggests stability increases with diversity because species respond dissimilarly to environmental conditions – species A has highest growth rates under conditions X whereas species B has highest growth rates under conditions Y. 
More species confer temporal stability by broadening the range of conditions under which the community maintains function [@Loreau2010].
Diverse models all predict a positive relationship between species richness and ecosystem stability [@Lehman2000; @Ives2002b; @Loreau2013], and experimental tests tend to support such a prediction [@Tilman2006; @Hector2010].

However, the ability of biodiversity-ecosystem functioning (BEF) experiments to represent real-world dynamics is the subject of debate [@Wardle2016; @Eisenhauer2016].
Much of the debate centers around the fact that species losses in BEF experiments are not allowed to be offset by species gains.
Theory typically suffers from the same limitation, in that it often ignores the processes that could lead to species gains in temporally variable environments [Loreau 2010, but see @Chesson2001].

Temporally fluctuating environmental conditions are an important ingredient for stable species coexistence, both in theoretical models [@Chesson2000; @Chesson2004] and in natural communities [@Caceres1997; @Descamps-Julien2005; @Adler2006; @Angert2009].
Such "fluctuation-dependent" coexistence [@Chesson2000] requires that species have unique environmental responses and that environmental conditions vary enough for each coexisting species to experience good and bad conditions.
Thus, there is reason to expect environmental variability to promote species richness when coexistence is maintained by a fluctuation-dependent mechanism [@Adler2008].
Of course, increasing environmental variability may also decrease ecosystem stability through time by increasing the fluctuations of individual species, regardless of species richness.

The countervailing effects of environmental variability present an interesting paradox: increasing variability should decrease ecosystem stability, but may also increase richness, which may offset the decrease in stability.
Such a paradox complicates predictions about how ecosystems will respond as environmental conditions exceed historical ranges of variability.
The unknown net effect of environment variability may be reflected in the mixed results from empirical studies on the diversity-stability relationship.
Observational tests of the diversity-stability relationship, which require sampling across natural diversity gradients, have yielded positive [@Hautier2014], neutral [@Valone2003; @Cusson2015], and negative [@Sasaki2011] relationships.
In a meta-analysis of diversity-stability relationships, @Jiang2009 found no significant evidence for an effect of species richness on ecosystem stability from observational studies in terrestrial ecosystems.
Thus, there appears to be a gap between the consistency of theoretical studies and the equivocation of empirical studies.

We argue this gap exists because the two bodies of theory that have developed to explain species coexistence on the one hand, and diversity-stability relationships on the other, have diverged.
One reason these two disciplines have diverged is because they have focused on slightly different questions.
Biodiversity-ecosystem stability studies typically ask how ecosystem variability responds to different levels of species richness at a given level of environmental variability [reviewed in @Kinzig2001; @Loreau2010], whereas coexistence studies ask how the long term stability of species coexistence responds to different levels of environmental variability [@Chesson1981a].

To reconcile these two bodies of theory, we extend theory on the relationship between species richness and ecosystem stability to cases in which species coexistence explicitly depends on environmental fluctuations and species-specific responses to environmental conditions.
We focus on the storage effect and relative nonlinearity using a general consumer-resource model.
We use the model to investigate two questions:

1. Does the direction of the diversity-stability relationship remain positive when species coexistence is fluctuation-dependent? And,

@@

@@

#### Implementing the Storage Effect
To make this a storage effect model, we need to satisfy three conditions: (1) the organisms must have a mechanism for persistence under unfavorable conditions, (2) species must respond differently to environmental conditions, and (3) the effects of competition on a species must be more strongly negative in good years relative to unfavorable years. 
Our model meets condition 1 because we include a dormant stage with very low death rates. 
We satisfy condition 2 with our model whenever $\gamma$ is not perfectly correlated between species. 
Lastly, our model meets condition 3 because condition 2 partitions intraspecific and interspecific competition into different years. 
Thus, during a high $\gamma$ year for one species, resource uptake is also inherently high for that species, which increases intraspecific competition relative to interspecific competition. 
So, given adequate variability in $\gamma$, the inferior competitor can persist. 
We created competitive hierarchies in the storage effect version of the model by altering species' biomass conversion efficiencies ($\epsilon$)

@@

#### Implementing Relative Nonlinearity
When considering consumer-resource dynamics, species coexistence by relative nonlinearity requires that each species has different nonlinear responses to resource availability, and resource availability must fluctuate through time.
In a constant resource environment, the species with the lowest $R^*$ will always exclude the other species. 
To create competitive hierarchies among species we manipulated species resource uptake curves such that the species with the lowest $R^*$ also had the lowest maximum growth rate at high resource levels (e.g., low *r*, low *a*, and low *b* values relative to other species in the Hill equation; Fig. SX).
Thus, our simulated species represent a continuum from resource-conservative to resource-acquisitive.
We still allow the germination rate ($\gamma$) to vary, but both species are perfectly correlated -- that is, $\rho=1$ (Fig. 1).

### Numerical simulations
To understand how fluctuation-dependent coexist can affect the diversity-stability relationship, we simulated the model with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many species can coexist, akin to a community assembly experiment with a species pool of four species.
This required simulating communities with all species initially present across a gradient of annual resource variability (for relative nonlinearity) or environmental cue variability (for the storage effect).
Second, we chose parameter values that allowed coexistence of all four species and performed species removals.
The two simulation experiments correspond to (i) sampling ecosystem function across a natural gradient of species richness and (ii) sampling ecosystem function across diversity treatments within a site.

To understand how increasing environmental variability will impact ecosystem stability when coexistence is fluctuation-dependent, we simulated the model over a range of environmental cue variability and species pool sizes.
Thus, for each size of species pool (1, 2, 3, or 4 species), we simulated the model at 15 evenly-spaced levels of environmental cue (range = 0.1,2) for the storage effect and 25 evenly-spaced levels of resource variability (range = 0.1,1.4) for relative nonlinearity.
We also explored the influence of species asymmetries in competitive ability and species' correlations of environmental responses in the storage effect model.
Under the storage effect, if all species are perfectly symmetrical, that is, there is no superior competitor, then coexistence is fluctuation independent.
We use one such parameterization of our model to contrast the response of ecosystem stability in fluctuation-dependent and fluctation-independent communities to environmental variation.
Likewise, under relative nonlinearity, species' resource response curves (Fig. SX) reflect traits that determine the intrinsic stability of each species.
Therefore, we ran two sets of simulations for relative nonlinearity: one where the species pools increased from stable to unstable species and vice versa.
For example, if species A is the most stable species and species D is the least stable we ran simulations where the species pool increased from one to four species as A then B then C then D.
We then ran simulations with that order reversed.

All simulations were run for 5,000 seasons with 20-day growing seasons. 
We averaged biomass over the growing season, and those yearly values were used to calculate total community biomass in each year.
After discarding an initial 500 seasons to reduce transient effects on our results, we calculated the coffecient of variation (*CV*) of summed species biomass through time.
Therefore, in our results we refer to ecosystem variability, which is the inverse of ecosystem stability.
We calculated species richness as the number of species whose average biomass was greater than 1 over the course of the simulation.
Parameter values for specific results are given in figure captions.
Within-season dynamics were solved given initial conditions using the package ``deSolve`` [@Soetaert2010] in ``R`` [@Team2013].
All model code has been deposited on Figshare (*link*) and is available on GitHub at http://github.com/atredennick/Coexistence-Stability.

RESULTS AND DISCUSSION
----------------------
### The diversity-variability relationship
The direction of the diversity-variability relationship can be positive and negative when species coexistence is maintained by fluctuation-dependent mechanisms (Fig. 2).
Ecosystem variability is positively correlated with species richness when species richness is measured across a gradient of environmental variability, which maintains diversity and promotes ecosystem variability (Fig. 2A,C).
If environmental conditions are sufficient to maintain coexistence, removing species increases ecosystem variability (Fig. 2B,D).
Thus, our results both confirm and contrast with theoretical and experimental findings that diversity begets stability.

When we held environmental variability constant and removed species, we produced the typical negative diversity-variability relationship (Fig. 2B,D), consistent with theoretical expecations from models with species coexistence maintained by fluctuation-dependent mechanisms.
Likewise, our results from the species removal simulations are consistent with results from biodiversity-ecosystem functioning experiments showing a negative relationship between species richness and ecosystem variability.
This is encouraging because species almost certainly coexist by some combination of fluctuation-independent (e.g., resource partitioning) and fluctuation-dependent mechanisms.
By extending theory to communities where species richness is explicitly maintained by temporal variability, we have gained confidence that experimental findings are generalizable to many communities.
In other words, in local settings where environmental variability is relatively homogenous, reductions in the number of species will reduce the stability of ecosystem functioning, regardless of how coexistence is maintained.

When we allowed a gradient of environmental variability to determine species coexistence, we discovered a positive relationship between species richness and ecosystem variability (Fig. 2A,C).
While surprising when viewed through the lens of previous theory and experimental findings, such a relationship is a direct consequence of how diversity can be maintained in fluctuating environments.
The storage effect and relative nonlinearity both require envrionmental fluctuations to allow niche differentiation between species pairs [@Chesson2000].
Therefore, species coexistence gains strength, for both mechanisms, as the environment becomes more variable (Fig. SX).

>>>>

\begin{abstract}
Theory relating species richness to ecosystem stability typically ignores the potential for environmental variability to promote species coexistence.
Failure to account for fluctuation-dependent coexistence mechanisms may explain observed deviations from the expected positive diversity-stability relationship, and limits our ability to predict the consequences of increasing environmental variability.
We use a consumer-resource model to explore how coexistence via the temporal storage effect and relative nonlinearity affects ecosystem stability.
We show that a negative, rather than positive, diversity-stability relationship is possible when ecosystem function is sampled across a natural gradient in environmental variability and diversity.
We also show how fluctuation-dependent coexistence can buffer ecosystem functioning against increasing environmental variability by allowing more species to coexist and contribute to portfolio effects.
Our work provides a general explanation for variation in observed diversity-stability relalationships and highlights the importance of conserving regional species pools to help buffer ecosystems against future increases in environmental variability.
\vspace{2em}
\end{abstract}

@@

INTRODUCTION
------------
@MacArthur1955, @Elton1958, and even Darwin [@Turnbull2013] recognized that species can compensate for each other and stabilize ecosystem functioning in fluctuating environments. 
This idea underlies the "insurance hypothesis" [@Yachi1999], which suggests stability increases with diversity because species respond dissimilarly to environmental conditions, broadening the range of conditions under which the community maintains function [@Loreau2010].
Diverse models all predict a positive relationship between species richness and ecosystem stability [@Lehman2000; @Ives2002b; @Loreau2013], and experimental tests tend to support such a prediction [@Tilman2006; @Hector2010].

However, the ability of biodiversity-ecosystem functioning (BEF) experiments to represent real-world dynamics is the subject of debate [@Wardle2016; @Eisenhauer2016].
Much of the debate centers around the fact that species losses in BEF experiments are not allowed to be offset by species gains.
Theoretical work on diveristy-stability relationships typically suffers from the same limitation, ignoring the processes by which temporal environmental variability can promote coexistence and increase species richness [Loreau 2010, but see @Chesson2001].

Fluctuating environmental conditions are an important ingredient for stable species coexistence, both in theoretical models [@Chesson2000; @Chesson2004] and in natural communities [@Caceres1997; @Descamps-Julien2005; @Adler2006; @Angert2009].
Such "fluctuation-dependent" coexistence requires that species have unique environmental responses and that environmental conditions vary so that each species in a community experiences favorable and unfavorable conditions, which prevents competitive exclusion [@Chesson2000].
When coexistence is maintained by a fluctuation-dependent mechanism, an increase in environmental variability might lead to an increase in species richness and, consequently, an increase in ecosystem stability.
However, increasing environmental variability may also decrease ecosystem stability by increasing the fluctuations of individual species, regardless of species richness.

The countervailing effects of environmental variability present an interesting paradox: increasing variability should decrease ecosystem stability, but the potential for an increase in richness might offset the decrease in stability.
Such a paradox complicates predictions about how ecosystems will respond as environmental conditions exceed historical ranges of variability.
The unknown net effect of environment variability may be reflected in the mixed results from empirical studies on the diversity-stability relationship.
Observational tests of the diversity-stability relationship, which require sampling across natural diversity gradients, have yielded positive [@Hautier2014], neutral [@Valone2003; @Cusson2015], and negative [@Sasaki2011] relationships.
In a meta-analysis of diversity-stability relationships based on observational studies in terrestrial ecosystems, @Jiang2009 found no significant evidence for an effect of species richness on ecosystem stability.
The idiosyncratic results of these empirical studies contrasts with the consistent conclusions from theoretical work.

The gap between theoretical expectations and empirical results of richness-stability relationships reflects the divergence of theory developed to explain species coexistence and theory developed to explain diversity and stability.
One reason these two disciplines have diverged is because they have focused on slightly different questions.
Diversity-stability studies typically ask how ecosystem variability responds to different levels of species richness at a given level of environmental variability [reviewed in @Kinzig2001; @Loreau2010], whereas coexistence studies ask how the long term stability of species coexistence responds to different levels of environmental variability [@Chesson1981a].

To reconcile these two perspectives, we extend theory on the relationship between species richness and ecosystem stability to cases in which species coexistence explicitly depends on environmental fluctuations and species-specific responses to environmental conditions.
We focus on the temporal storage effect and relative nonlinearity using a general consumer-resource model.
We use the model to investigate three questions:

1. Does the direction of the diversity-stability relationship remain positive when species coexistence is fluctuation-dependent?

@@

3. Do our answers to the previous two questions depend on the specific fluctuation-dependent coexistence mechanism (i.e., storage effect vs. relative nonlinearity)?

@@

#### Implementing the Storage Effect
For the storage effect to operate, we need species-specific responses to environmental variability, positive environment-competition (EC) covariance, and subadditive population growth (@Chesson1994; @Chesson2000a). 
EC covariance limits population growth of residents: in a favorable year, they are limited by strong intraspecific competition because their populations are large. 
However, a species at very low abundance can escape the limitations of EC covariance if it experiences favorable conditions when the residents it is competiting against experience (relatively) unfavorable conditions. 
Subadditive population growth buffers populations against large population decreases in unfavorable years.

Our model includes subadditive population growth because we include a dormant stage with very low death rates, which limits large population declines.
Our model includes EC covariance through the effect of the dormant-to-live transition rate, $\gamma$, on population growth.
A high $\gamma$ year for a species represents a good *E* year in which live biomass, and thus resource uptake, is above average. 
These conditions correspond to *C* also being above average because a common species, by definition, cannot escape intraspecific competition.

In our model, rare species are rare because they are inferior competitors. 
So, given adequate variability in $\gamma$, the inferior competitor can persist because its rarity allows it to take advantage of favorable envrionmental conditions. 
We created competitive hierarchies in the storage effect version of the model by altering species' biomass conversion efficiencies ($\epsilon$)

@@

#### Implementing Relative Nonlinearity
\textcolor{red}{YOU NEED TO REMIND READERS ABOUT THE STANDARD RESOURCE COMPETITION MODELS, MACARTHUR, TILMAN, ETC. IN GENERAL, I THINK THIS PARAGRAPH NEEDS TO BE EXPANDED AND EXPLAINED MORE SLOWLY, STEP BY STEP.}

When considering consumer-resource dynamics, species coexistence by relative nonlinearity requires that each species has different nonlinear responses to resource availability, and resource availability must fluctuate through time.
In a constant resource environment, the species with the lowest $R^*$ will always exclude the other species. 
To create competitive hierarchies among species we manipulated species resource uptake curves such that the species with the lowest $R^*$ also had the lowest maximum growth rate at high resource levels (e.g., low *r*, low *a*, and low *b* values relative to other species in the Hill equation; Fig. SX).
Thus, our simulated species represent a continuum from resource-conservative to resource-acquisitive.
We still allow the germination rate ($\gamma$) to vary, but both species are perfectly correlated -- that is, $\rho=1$ (Fig. 1).

### Numerical simulations
To explore how fluctuation-dependent coexistence can affect the diversity-stability relationship, we simulated the model with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many species can coexist, akin to a community assembly experiment with a species pool of four species.
To do this, we simulated communities with all species initially present across a gradient of annual resource variability (for relative nonlinearity) or environmental cue variability (for the storage effect).
Second, we chose parameter values that allowed coexistence of all four species and then performed species removals, akin to a biodiversity-ecosystem function experiment.
The two simulation experiments correspond to (i) sampling ecosystem function across a natural gradient of species richness and (ii) sampling ecosystem function across diversity treatments within a site.

To understand how increasing environmental variability will impact ecosystem stability when coexistence is fluctuation-dependent, we simulated the model over a range of species pool sizes and environmental cue or resource variability.
For each size of species pool (1, 2, 3, or 4 species), we simulated the model at 15 evenly-spaced levels of environmental cue (range = 0.1,2) for the storage effect and 25 evenly-spaced levels of resource variability (range = 0.1,1.4) for relative nonlinearity.
We also explored the influence of asymmetries in species' competitive abilities and correlations in species' environmental responses within the storage effect model.

Under relative nonlinearity, species' resource response curves (Fig. SX) reflect traits that determine the intrinsic stability of each species.
Therefore, we ran two sets of simulations for relative nonlinearity: one where the species pools increased from stable to unstable species and vice versa.
For example, if species A is the most stable species and species D is the least stable we ran simulations where the species pool increased from one to four species as A then B then C then D.
We then ran simulations with that order reversed.

All simulations were run for 5,000 seasons with 20-day growing seasons. 
We averaged biomass over the growing season, and those yearly values were used to calculate total community biomass in each year.
After discarding an initial 500 seasons to reduce transient effects on our results, we calculated the coffecient of variation (*CV*) of summed species biomass through time, which represents ecosystem variability,  the inverse of ecosystem stability.
We calculated species richness as the number of species whose average biomass was greater than 1 over the course of the simulation.
Parameter values for specific results are given in figure captions.
Within-season dynamics were solved given initial conditions using the package ``deSolve`` [@Soetaert2010] in ``R`` [@Team2013].
All model code has been deposited on Figshare (*link*) and is available on GitHub at http://github.com/atredennick/Coexistence-Stability.

RESULTS
-------
Our simulation experiments show that the direction of the diversity-variability relationship can be positive or negative when species coexistence is maintained by fluctuation-dependent mechanisms (Fig. 2).
Ecosystem variability is positively correlated with species richness when species richness is measured across a gradient of environmental variability. 
The variability promotes diversity (Fig. SX) but also leads to increased ecosystem variability (Fig. 2A,C).
On the other hand, holding environmental conditions constant (but at levels sufficient for coexistence), removing species increases ecosystem variability (Fig. 2B,D).

DISCUSSION
----------
### The diversity-variability relationship
Theory developed for biodiversity-ecosystem function experiments emphasizes that increases in species richness should reduce ecosystem variability. , consistent with theoretical expectations from models in which species coexistence maintained by fluctuation-dependent mechanisms, and consistent with Species almost certainly coexist by some combination of fluctuation-independent (e.g., resource partitioning) and fluctuation-dependent mechanisms.
By extending theory to communities where species richness is explicitly maintained by temporal variability, we have gained confidence that experimental findings are generalizable to many communities.
In other words, in local settings where environmental variability is relatively homogenous, reductions in the number of species will reduce the stability of ecosystem functioning, regardless of how coexistence is maintained.
